\myfont Implementation and evaluation of register allocation for translating LLVM-{}- to x86-64

### 2024-01-11 article.tex
Add spilling heuristics example
Elaborate run time of lva

\noindent This is unnoticeable for snippets of marginal size like most examples included, but when applied to functions with a significant amount of def-use paths, it can be quite consequential. For \texttt{tests/sha256.ll} for instance, the running time of this analysis is effectively doubled:

\begin{minted}{bash}
$ time dune exec build -- tests/sha256.ll  -t lva -r
33,69s user 0,04s system 99% cpu 33,765 total
$ time dune exec build -- tests/sha256.ll  -t lva
17,26s user 0,05s system 99% cpu 17,335 total
\end{minted}

@@

\section{Linear Scan}
Another approach that also depends on liveness analysis is the linear scan algorithm.
It is a simple and efficient approach that performs a linear pass over IR to produce a register assignment significantly faster than all of the steps involved with graph coloring.
That said, it may generate more imprecise assignments and therefore also be less performant at runtime, so the utility of such an approach depends on context.
Some compilers typically compile code into machine code ahead of time (AOT) while others depend on just-in-time (JIT) compilation which defers the compilation of code until runtime.
Therefore linear scan is an approach to register allocation that may be used to achieve reasonably performant runtime compilation.
%Whether or not to  a cost depends on the context, as JIT compilation will often 
%reduced compile-time in  the fact that assignment is fast is a benefit in some cases.

The primary text is \textit{Linear Scan Register Allocation} \cite{linear}, the application of which also relies on prior liveness analysis. Given liveranges, assignments are easily derived from only one pass through IR.
Similar to graph coloring, given \(k\) available GPRs, it assigns  as many variables as possible, and spilling the rest.

It does this by first building the intervals using the \(\mathit{in}[s]\) and \(\mathit{out}[s]\) resulting from prior dataflow analysis. This can be done in a single pass with a little creative array indexing as can be seen in Figure \ref{fig:intervals-ocaml}.

>>>>

\noindent This is unnoticeable for snippets of marginal size like most examples included, but when applied to functions with a significant amount of def-use paths, it can be quite consequential. For \texttt{tests/sha256.ll} for instance, the running time of this analysis is effectively doubled (see Figure \ref{fig:lva-time}).

\begin{figure}[H]
  \centering
  \begin{minipage}[b]{0.69\textwidth}
    \begin{minted}[bgcolor=LightGray]{bash}
$ time dune exec build -- tests/sha256.ll -t lva -r
36.91s user 0,04s system 99% cpu 33,765 total
$ time dune exec build -- tests/sha256.ll -t lva
18,61s user 0,03s system 99% cpu 19,055 total
    \end{minted}
    \caption{Average runtime over several runs (n=32)}\label{fig:lva-time}
  \end{minipage}
\end{figure}

@@

\section{Linear Scan}
Another approach that depends on liveness analysis is the linear scan algorithm.
It is a simple and efficient approach that performs a linear pass over IR to produce a register assignment significantly faster than the steps involved with graph coloring.
That said, it may generate more imprecise assignments and therefore also be less performant at runtime, so the utility of such an approach depends on the context in which it is used.
While compilers like clang translate to machine code ahead-of-time (AOT), languages like C\# and Java will often rely on just-in-time (JIT) compilation to translate their respective VM instructions to native machine code,

Such VMs that depend on JIT will often defer compilation to runtime. This is done for a myriad of reasons, portability and security among them, but usually at the cost of runtime performance. One of the reasons for this can be attributed to the way in which registers are allocated, as a significant downside of JIT is the fact it needs to be compiled in full before execution can begin. Therefore, generally a faster class of allocators is employed, linear scan being one of them, as linear scan is a reasonably performant alternative to graph coloring, that may be used to achieve reasonably efficient register assignments at runtime.
%Whether or not to  a cost depends on the context, as JIT compilation will often 
%reduced compile-time in  the fact that assignment is fast is a benefit in some cases.

The primary text is \textit{Linear Scan Register Allocation} \cite{linear}, the application of which also relies on prior liveness analysis. Given liveranges, assignments are easily derived from only one pass through IR.
Similar to graph coloring, given \(k\) available GPRs, it assigns  as many variables as possible, and spills the rest.

It does this by first building the intervals using the \(\mathit{in}[s]\) and \(\mathit{out}[s]\) resulting from prior dataflow analysis.
This can be done in a single pass with some creative array indexing, but my approach relies on two passes: one for finding the start points (see Figure \ref{fig:livenessstarts.ml}), the other for finding the end points.

\begin{figure}[H]
  \centering
   \begin{minipage}[b]{0.88\textwidth}
     \centering
     \begin{minted}[linenos,bgcolor=LightGray]{ocaml}
let add e = function
  | Some s -> Some (S.SS.add e s)
  | None -> Some (S.SS.singleton e)

let intervalstart insns (in_, out) =
  let insn (ordstarts, starts, active) (i, _n) =
    let out' = S.SS.diff (S.SS.union in_.(i) out.(i)) active in
    ( S.SS.fold (fun e a -> IT.update i (add e) a) out' ordstarts,
      S.SS.fold (fun e a -> S.ST.add e i a) out' starts,
      S.SS.union active out' )
  in
  List.fold_left insn (IT.empty, S.ST.empty, S.SS.empty) insns
     \end{minted}
     \caption{Implementation of \texttt{intervalstart} in \texttt{lib/linear.ml}}\label{fig:livenessstarts.ml}
   \end{minipage}
\end{figure}

\noindent Of note are the structures over which the folding is done: the \texttt{IT.t} is a \texttt{Map.S} module indexing earlier described symbol sets (i.e. \texttt{S.SS.t}). The reason for this is to iterate through intervals in increasing and decreasing order of start/length and end points respectively. The symbol table (\texttt{S.ST.t}) is to allow for reverse lookup, i.e. which program point variable \(s\) is defined, and lastly the symbol set (i.e. \texttt{S.SS.t}) which is just there to prevent multiple start points for the same variable.

\begin{figure}[H]
  \centering
   \begin{minipage}[b]{0.95\textwidth}
     \centering
     \begin{minted}[linenos,bgcolor=LightGray]{ocaml}
let intervalends insns starts (in_, out) =
  let insn (i, _n) (ordstarts, lengths, ends, active) =
    let in' = S.SS.diff (S.SS.union in_.(i) out.(i)) active in
    ( S.SS.fold (fun e a -> IT.update i (add e) a) in' ordstarts,
      S.SS.fold
        (fun e a -> IT.update (-i + S.ST.find e starts) (add e) a)
        in' lengths,
      S.SS.fold (fun e a -> S.ST.add e i a) in' ends,
      S.SS.union active in' )
  in
  List.fold_right insn insns (IT.empty, IT.empty, S.ST.empty, S.SS.empty)
     \end{minted}
     \caption{Implementation of \texttt{intervalendsintervalstart} in \texttt{lib/linear.ml}}\label{fig:intervalends.ml}
   \end{minipage}
\end{figure}

\noindent The opposite is quite similar (see Figure \ref{fig:intervalends.ml}).

as can be seen in Figure \ref{fig:intervals-ocaml}.

\subsection{Heuristics spilling}
\usetikzlibrary{chains,arrows}
\begin{figure}[H]

     \centering
     \begin{minipage}[b]{0.30\textwidth}
     \centering

     \begin{tikzpicture}[scale=0.442,
    node distance=0mm, text height=1.5ex, text depth=.25ex,
    attribute/.style={rectangle,minimum size=8mm,very thin,draw=black!50,font=\ttfamily},
    link/.style={thin, to path={-- ++(0,8mm) -| (\tikztotarget)}, >= triangle 45, shorten >= 0pt}]

%\node (n1) [attribute] {\small NODE1};
%\node (n2) [attribute, right=of n1] {\small NODE2};

   \node at (7,0) (11) {};
   \node at (7,1) (10) {};
   \node at (7,2) (9) {};
   \node at (7,3) (8) {};
   \node at (7,4) (7) {};
   \node at (7,5) (6) {};
   \node at (7,6) (5) {};
   \node at (7,7) (4) {};
   \node at (7,8) (3) {};
   \node at (7,9) (2) {};

   \draw[|-|] (2)  node[above,transform canvas={xshift=1.43cm}] {\texttt{\%3}} edge [transform canvas={xshift=1.5cm}] (6);
   \draw[|-|] (3) node[above,transform canvas={xshift=0.83cm}] {\texttt{\%4}} edge [transform canvas={xshift=0.9cm}] (8);
   \draw[|-|] (4) node[above] {\texttt{\%5}} edge [ ]  (11);

   %\draw[help lines] (0,0) grid (3,11);
   %\node at (7,0) (11) {};
   %\node at (7,1) (10) {};
   %\node at (7,2) (9) {};
   %\node at (7,3) (8) {};
   %\node at (7,4) (7) {};
   %\node at (7,5) (6) {};
   %\node at (7,6) (5) {};
   %\node at (7,7) (4) {};
   %\node at (7,8) (3) {};
   %\node at (7,9) (2) {};
   %\path (2.center) edge node[above] {} (4.center);
   %\path (5.center) edge node[above] {} (7.center);
   %\draw[|-|] (8.center)  edge [transform canvas={xshift=3mm]]   (10.center);

\end{tikzpicture}
\vspace{0.7em}
     \caption{Live ranges}\label{fig:linear.lra}
   \end{minipage}
   \begin{minipage}[b]{0.68\textwidth}
     \centering
     \begin{minted}[linenos,bgcolor=LightGray]{llvm}
define i32 @main (i32 %0, i8* %1) {
  %3 = getelementptr i8*, i8** %1, i64 1
  %4 = load i8*, i8** %3
  %5 = call i32 @atoi (i8* %4)
  call i32 @printf (i8* @.str, i8** %3, i8* %4, ...
  %7 = call i32 @isqrt (i32 %5)
  %8 = call i64 @strtoll (i8* %4, i8* null, i32 10)
  call i32 @printf (i8* @.str.1, i32 %5, i64 %8)
  call i32 @printf (i8* @.str.2, i32 %7, i32 %5)
  ret i32 %5
}
     \end{minted}
     \caption{\texttt{@main} of \texttt{tests/linear.ll}}\label{fig:linear.ll}
   \end{minipage}
\end{figure}
